---
title: "Data checker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data checker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`data.checker` is a package for helping with boilerplate data checks. It enables you to automate fundamental data checks which, while simple, can be time-consuming to implement.

`data.checker`

-   Checks data against a user supplied schema that defines what columns and data types are expected

-   Enables user to add additional custom data checks based on multiple columns

-   Creates exports of the results for QA

## The basics

Initialising the data checker is simple. All you need to supply are a dataset and a schema. The schema is a named list that tells the data checker what sorts of columns and values to expect.

### Example dataset:

```{r}
library(data.checker)

df <- data.frame(
  age = c(10, 11, 13, 15, 22, 34, 80),
  sex = c("M", "F", "M", "F", "M", "F", "M")
)

df
```

### Example schema:

```{r}
schema <- list(
  check_duplicates = FALSE,
  check_completeness = FALSE,
  columns = list(
    age = list(type = "integer", optional = FALSE),
    sex = list(type = "character", optional = FALSE)
  )
)

schema
```

### Initialising the data checker

Running the `new_validator` function will create a `Validator` object.



```{r}
validator <- data.checker::new_validator(
  data = df,
  schema = schema
)
```

```{r echo = FALSE}
anonymise_validator <- function(validator_object){
  
  validator_object[["log"]][[1]][["description"]] <-  "Date: 2025-01-01\nsysname: Windows\nrelease: 10 x64\nversion: \nnodename: \nmachine: \nlogin: username\nuser: username\neffective_user: username\nudomain: \nR version : R version 4.5.1 (2025-06-13 ucrt)\ndata.checker version: 0.0.0.9000\n"
  
  return(validator_object)
  
}

validator <- anonymise_validator(validator)
```

```{r}
print(validator)
```

The `Validator` object logs system information which can be exported along with the QA log, meaning you have a comprehensive record of what was done, when and on what systems. Printing the `Validator` object will show you the current QA log.

### Running checks

The `check` function will run the full suite of checks on your `Validator` object and add them to the log.

```{r}
check_results <- data.checker::check(validator)

print(check_results)
```

### Exporting your log

The `export` function will export your log in html, csv, yaml or json. We strongly recommend exporting these automated QA logs along with your outputs so you have a record of which automated checks were done and what they found.

```{r eval = FALSE}
data.checker::export(check_results, file = "example.html", format = "html")
```

Alternatively, you can use the `validate` function to run the full process.

```{r eval = FALSE}
data.checker::check_and_export(df, schema, file = "example.html", format = "html", hard_check = FALSE)
```

## Setting up the schema

The schema has certain mandatory and optional fields.

### Mandatory fields

`check_duplicates`: TRUE or FALSE. If TRUE, the dataset will be checked for duplicate rows. `check_completeness`: TRUE or FALSE. If TRUE, the dataset will be checked to ensure there is at least one row for all combinations of factors. `columns`: a list of column names with an entry for each column.

For each column, you should include a type ("character", "integer", "double", "logical"). You also need an "optional" setting (TRUE or FALSE) if TRUE the checker will raise an error if the column is missing. If FALSE the checker data will not raise an error if it's missing.

You can also optionally define a class if you want it to be checked. There are three special types you can choose - "Date", "datetime" and "factor". In R, these are implemented as a specific combination of types and classes, but the data checker simplifies this for you by setting up those parts of the schema for you.

### Optional checks:

Optional checks can be applied to each column depending on the column type. In the scheme, these should form part of the `columns` list.

-   all types:
    -   allow_na (TRUE/FALSE): checks if there are any missing values
    -   class (character vector of any length): checks class of column
-   integer/double checks:
    -   min_val (numeric): minimum value
    -   max_val (numeric): maximum value
    -   min_decimal (double only, numeric): minimum number of decimal places
    -   max_decimal (double only, numeric): maximum number of decimal places
-   character checks:
    -   min_length (numeric): minimum number of characters
    -   max_length (numeric): maximum number of characters
    -   allowed strings (character): either a list of allowed strings or a regular expression (see regular expression guide below)
    -   forbidden_strings (character): either a list of forbidden strings or a regular expression
-   Date checks:
    -   min_date (character): minimum date using the format "YYYY-MM-DD", e.g. "2025-08-19"
    -   max_date (character): maximum date using the format "YYYY-MM-DD", e.g. "2025-08-19"
-   datetime checks:
    -   min_datetime (character): minimum time using the format "YYYY-MM-DD HH-MM-SS". Datetime is more flexible than Date, meaning the hour, minute and second parts are optional. It will accept the formats: Y, YM, YMD, YMDH, YMDHM and YMDHMS.
    -   max_datetime (character): same format as min_datetime

### Loading from file

Schema objects can get pretty large and clutter your code. You should also avoid needing to edit your code every time you want to change your schema. Instead, you can create your schema as a yaml or json file instead. You can then supply `new_validator` with the file path, and the package will do the rest.

```{r}
df <- data.frame(
  id = 1:10,
  age = c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
  sex = c("M", "F", "M", "F", "M", "F", "M", "F", "M", "F"),
  country = factor(
    c("England", "England", "Wales", "Scotland", "Wales", "England", "Northern Ireland", "Wales", "Scotland", "Northern Ireland"),
    levels = c("England", "Scotland", "Wales", "Northern Ireland")),
  date = lubridate::ymd(c(
    "2021-01-01",
    "2021-02-01",
    "2021-02-01",
    "2021-03-01",
    "2021-03-01",
    "2021-03-01",
    "2021-04-01",
    "2021-04-01",
    "2021-04-01",
    "2021-05-01"
  ))
)

data_check_results <- data.checker::new_validator(schema = "example_schema.yaml", data = df) |> 
  data.checker::check()
```

```{r echo = FALSE}
data_check_results <- anonymise_validator(data_check_results)
```

```{r}
print(data_check_results)
```

### Custom checks

You can write your own checks using the `add_custom_check` function. This is particularly useful for checks involving more than one column, which cannot be configured using the standard template. The checks are done in the context of the original data, meaning you can reference columns as if they are variables in the environment (similar to tidy evaluation). This is recommended because it guarantees the checks are done on the correct data only. Alternatively, you can use standard evaluation (see example below).

```{r}
df <- data.frame(
  id = 1:10,
  age = c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
  sex = c("M", "F", "M", "F", "M", "F", "M", "F", "M", "F")
)

schema <- list(
  check_duplicates = FALSE,
  check_completeness = FALSE,
  columns = list(
    id = list(type = "double", optional = FALSE),
    age = list(type = "double", optional = FALSE),
    sex = list(type = "character", optional = FALSE)
  )
)

data_check_results <- data.checker::new_validator(df, schema) |>
  data.checker::check() |>
  data.checker::add_check(description = "There are no males over 90 (tidy evaluation)", condition = !(sex == "M" & age > 90)) |>
  data.checker::add_check(description = "There are no males over 90 (standard evaluation)", condition = !(df$sex == "M" & df$age > 90))


```

```{r}
data_check_results <- anonymise_validator(data_check_results)
```

```{r}
print(data_check_results)
```

### Custom log entries

You can choose to add your own entries to the QA log using the `add_qa_entry` function. The function expects a datachecker object as the first argument and a description. You can also optionally add:

-   `failing_ids`: a vector containing the columns/rows that failed the checks
-   `outcome`: TRUE/FALSE for passing/failing checks or NA if you want to leave the field blank. Defaults to NA
-   `entry_type`: either "info", "warning" or "error". Info = neutral log record, warning = something is wrong but could be safely ignored, error something is wrong that is likely to break your code. Defaults to "info".

```{r}
df <- data.frame(
  age = c(10, 11, 13, 15, 22, 34, 80),
  sex = c("M", "F", "M", "F", "M", "F", "M")
)

schema <- list(
  check_completeness = FALSE,
  check_duplicates = FALSE,
  columns = list(
    age = list(type = "integer", optional = FALSE),
    sex = list(type = "character", optional = FALSE)
  )
)

validator <- data.checker::new_validator(df, schema)

validator <- data.checker::add_qa_entry(
  validator, 
  description = "Example custom log entry",
  entry_type = "info"
)

```

```{r}
validator <- anonymise_validator(validator)
```

```{r}
print(validator)
```

